# (PART\*) Inferential Statistics {-}

# Introduction {#inferential_stats}

>**Learning Objectives** 
To develop and apply population parameter point estimates  
To create and interpret confidence intervals  
To understand p-values and statistical significance  
To test hypotheses using statistical techniques

By conducting a statistical analysis, we usually want to *infer* something about the population under study by applying statistical methods to the data collected. These inferences fall into two categories: estimation and hypothesis testing.

## Estimation

Estimation is the process of inferring the value of a population parameter from a sample. This involves deriving point estimates and confidence intervals.

As an example, imagine that a researcher records the mass in kilograms of 100 Yellowfin Bream (Acanthopagrus australis) caught along the NSW coast. The sample has a mean of 1.3kg and a standard deviation of 0.4kg.

>**Point Estimate** - A single-valued statistic given as an estimate for a population parameter e.g. the mean of a sample ($\bar{x}$) is given as the estimate for the population mean ($\mu$)

The point estimate for the Bream example would be the sample mean (1.3kg).

>**Confidence Interval (CI)** - There is an n% probability that the parameter of interest is contained within a given confidence interval. Hence, for a 95% CI, there is a probability of 95% that the interval contains the parameter. 

A 95% CI for the mean of the Bream example would be $1.3\pm1.96\times\frac{0.4}{\sqrt{100}} = [1.22,1.38]$, assuming that the sample is from a normally distributed population. There is a 95% probability that this interval covers the true population mean.

>The formula for calculating confidence intervals depends on the parameter of interest. Statistical software can calculate automatically the most common confidence intervals for a given data set.


## Hypothesis Testing

The boxplot below suggests there is a difference in the mean between the trt1 and trt2 groups, but how can we test this hypotheses in a quantitative way? This is the purpose of statistical tests.


```{r}
PlantGrowth %>% 
  ggplot(aes(x=group, y=weight)) +
  geom_boxplot()
```

The process for applying a statistical test can be summarised as:

1. Form a null hypothesis, $H_0$
2. Form an alternative hypothesis, $H_a$
3. Decide on an acceptable significance level (Type I error rate), $\alpha$, (usually 5%, or $\alpha = 0.05$)
3. Apply an appropriate statistical test
4. Use the p-value returned from the test to decide if the null hypothesis should be rejected
5. Report your results

Continuing with the example above,

$H_0:$ There is no difference in the means of trt1 and trt2.  
$H_a:$ There is a statistically significant difference in the means of trt1 and trt2

The appropriate test for the difference in means for two independent groups is the **independent t-test**. Applying this test with a **95% confidence level** to the two groups gives $p=0.009298$.

The results of the test could be reported as

"Applying the independent samples t-test gave a p-value below the significance level ($\alpha = 0.05, p = 0.009298$). Hence, the null hypothesis was rejected."

>Notice that the null hypothesis was rejected. It would be incorrect to state that the alternative hypothesis was accepted because at the stated significance level, there was still a 5% chance that the null hypothesis was correct.

```{r results='hide'}
x = PlantGrowth %>% filter(group == "trt1") %>% select(weight)
y = PlantGrowth %>% filter(group == "trt2") %>% select(weight)

t.test(x, y, alternative = "two.sided", conf.level = 0.95)
```


* Effects, powers (detecting effects), and sample size
* Null and alternative hypotheses - aim to reject H0; example
* Review Type I/II errors


## Further Reading

[**Confidence Intervals, Pennsylvania State University**](https://online.stat.psu.edu/stat200/lesson/4)

[Estimation, Shafer and Zhang](https://stats.libretexts.org/Bookshelves/Introductory_Statistics/Introductory_Statistics_(Shafer_and_Zhang)/07%3A_Estimation)

[**Estimating a Confidence Interval, Navarro, D**](https://stats.libretexts.org/Bookshelves/Applied_Statistics/Learning_Statistics_with_R_-_A_tutorial_for_Psychology_Students_and_other_Beginners_(Navarro)/10%3A_Estimating_Unknown_Quantities_from_a_Sample/10.05%3A_Estimating_a_Confidence_Interval)
