# (PART\*) Inferential Statistics {-}

# Introduction {#inferential_stats}

>**Learning Objectives** 
To develop and apply population parameter point estimates  
To create and interpret confidence intervals  
To understand p-values and statistical significance  
To test hypotheses using statistical techniques

By conducting a statistical analysis, we usually want to *infer* something about the population under study by applying statistical methods to the data collected. These inferences fall into two categories: estimation and hypothesis testing.

## Estimation

Estimation is the process of inferring the value of a population parameter from a sample. This involves deriving point estimates and confidence intervals.

As an example, imagine that a researcher records the mass in kilograms of 100 Yellowfin Bream (Acanthopagrus australis) caught along the NSW coast. The sample has a mean of 1.3kg and a standard deviation of 0.4kg.

>**Point Estimate** - A single-valued statistic given as an estimate for a population parameter e.g. the mean of a sample ($\bar{x}$) is given as the estimate for the population mean ($\mu$)

The point estimate for the Bream example would be the sample mean, 1.3kg.

>**Confidence Interval (CI)** - There is an n% probability that the parameter of interest is contained within a given confidence interval. Hence, for a 95% CI, there is a probability of 95% that the interval contains the parameter. 

A confidence interval is calculated using the formula $\bar{x}\pm\text{MOE}$ where MOE is the *margin of error*.

For example, the MOE for the Bream example would be based on a t-distribution since the population standard deviation is unknown. This would give $\text{MOE}=1.9842\times\frac{0.4}{100}$, and the 95% CI for the mean would be $1.3\pm0.07937=[1.22, 1.38]$. There is a 95% probability that this interval covers the true population mean. 

>The method for calculating confidence intervals depends on the parameter of interest. Statistical software can calculate automatically the most common confidence intervals for a given data set.

>**Activity**  
Replicate the above calculations using the [Statistics Kingdom Confidence Interval Calculator](https://www.statskingdom.com/confidence-interval-calculator.html).


## Hypothesis Testing

The boxplot below suggests there is a difference in the mean between the trt1 and trt2 groups, but how can we test this hypotheses in a quantitative way? This is the purpose of statistical tests. 


```{r}
PlantGrowth %>% 
  ggplot(aes(x=group, y=weight)) +
  geom_boxplot()
```

A statistical test returns a **test statistic** which is used to decide if the data provides enough evidence to reject the null hypothesis. If the test statistic is beyond the predetermined **significance level** ($\alpha$), then the test result is described as **statistically significant**.

The process for applying a statistical test can be summarised as:

1. Form a null hypothesis, $H_0$
2. Form an alternative hypothesis, $H_a$
3. Decide on an acceptable significance level, $\alpha$, (usually 5%, or $\alpha = 0.05$)
3. Apply an appropriate statistical test
4. Use the p-value returned from the test to decide if the null hypothesis should be rejected
5. Report your results

Continuing with the example above,

$H_0:$ There is no difference in the means of trt1 and trt2.  
$H_a:$ There is a statistically significant difference in the means of trt1 and trt2

The appropriate test for the difference in means for two independent groups is the **independent t-test**. Applying this test with a 5% significance level ($\alpha = 0.05$) to the two groups gives $p = 0.009298$.

The results of the test could be reported as

"Applying the independent samples t-test produced a significant p-value ($p = 0.009298$) at the 5% significance level ($\alpha = 0.05$). Hence, the null hypothesis was rejected, and it was concluded that there was a statistically significant difference in the means of trt1 and trt2."

>Notice that the null hypothesis was rejected. It would be incorrect to state that the alternative hypothesis was accepted because at the stated significance level, there was still a 5% chance that the null hypothesis was correct.

Returning to the boxplot for the PlantGrowth data set, is it likely that there is a true difference in mean between the ctrl and trt1 groups? Perhaps the lower relative position of the trt1 boxplot is simply due to the randomness of sampling in this case, since there is considerable overlap between the two groups. Performing the same test as that used to compare groups trt1 and trt2 gives a non-significant p-value (0.2504), so there is not enough evidence to reject the null hypothesis. 

The results of the test could be reported as 

"Applying the independent samples t-test did not produce a significant p-value ($p = 0.2504$) at the 5% significance level ($\alpha = 0.05$). Hence, the null hypothesis was not rejected, and it was concluded that there was no statistically significant difference in the means of ctrl and trt1."


```{r results='hide'}
ctrl = PlantGrowth %>% filter(group == "ctrl") %>% select(weight)
trt1 = PlantGrowth %>% filter(group == "trt1") %>% select(weight)
trt2 = PlantGrowth %>% filter(group == "trt2") %>% select(weight)

t.test(trt1, trt2, alternative = "two.sided", conf.level = 0.95)
t.test(ctrl, trt1, alternative = "two.sided", conf.level = 0.95)
```
>**p-value Pitfalls**  
* Effects, powers (detecting effects), and sample size
* borderline p-values

* Review Type I/II errors


## Further Reading

[Estimation, Shafer and Zhang](https://stats.libretexts.org/Bookshelves/Introductory_Statistics/Introductory_Statistics_(Shafer_and_Zhang)/07%3A_Estimation)

[**Confidence Intervals, Pennsylvania State University**](https://online.stat.psu.edu/stat200/lesson/4)

[**Estimating a Confidence Interval, Navarro, D**](https://stats.libretexts.org/Bookshelves/Applied_Statistics/Learning_Statistics_with_R_-_A_tutorial_for_Psychology_Students_and_other_Beginners_(Navarro)/10%3A_Estimating_Unknown_Quantities_from_a_Sample/10.05%3A_Estimating_a_Confidence_Interval)



